---
title: "Copa de la Liga 2024 - Midfielder Scoring Calculator"
subtitle: "M9 Task - MSc Data Analytics in Football"
author: "Tommy Las"
date: "2024-04-28"
output: html_document
---
#Introduction

Using a scoring calculator algorithm, we rank the best performing midfielders in Copa de La Liga 2024 Argentina.
We import a CSV file with data from FBref, with only some modifications in columns names.
The goal for this task is to clean the data, select the KPIs that are suitable for a defending midfielder, and apply a scoring algorithm to each player.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Libraries

```{r libraries, warning=FALSE, message=FALSE, results='hide'}
library(readr)
library(tidyverse)
```

# Read CSV file

``` {r csv}
filename <- 'FBRef_2024_CopaDeLaLiga_MidfieldersAnalysis.csv'
raw_data <- read_csv(filename)
```

# Filter by position

We want to filter our data by position, we want players whose main position is *MF*.

``` {r filtering}
raw_data <- raw_data %>% filter(Pos == "MF")
```

We check the unique positions in our dataset.

``` {r}
cat("The unique positions in our data:", unique(raw_data$Pos[1]))
```

# Data filtering

Our next goal is to reduce our sample data. We want to filter players by *% minutes played*, *max age*. We are also selecting the features (metrics) that are specific for a defensive midfielder, reducing the number of columns.
We create the _filter_data_ function

``` {r data cleanup function}

filter_data <- function(data, metrics, max_age, min_minutes_percentage ){
  squad_matches_played <- 14 #copa de la liga matches before playoffs
  data_filter <- data %>%
    #Calculate the percentage a player played on their team, and compare to minimum percentage passed as parameter
    filter(min_minutes_percentage <= ((as.numeric(Min) / (90*squad_matches_played)) *100), 
           Age <= max_age) %>%
    select(c("Player", "Squad", "Min", metrics))
  
  return (data_filter)
}

```

```{r}
selected_metrics <- c("Progressive Passes", "Key Passes", "Interceptions", "Tackles" ,"Tackles Won", "Passes into Final Third", "Long Passes Attempted", "Long Passes Success %", "Total Passes Attempted", "Total Passes Success %","Total Blocks"  )
#Get our filtered dataset
filtered_data <- filter_data(raw_data, selected_metrics, 32, 30)

```


# Data Processing

Numeric metrics in our dataframe are strings, we need to modify them as numeric

```{r}

filtered_data[, 3:ncol(filtered_data)] <- sapply(filtered_data[, 3:ncol(filtered_data)], as.numeric)

```

There are some metrics in our dataframe that are total values, we want to know these metrics within a 90 minute frame, since one player may play more minutes than another player. We divide the metric by the 90s column.

```{r}

metrics_modified <- c("Progressive Passes", "Key Passes", "Interceptions", "Tackles", "Tackles Won","Passes into Final Third", "Long Passes Attempted", "Total Passes Attempted", "Total Passes Success %")

filtered_data[, metrics_modified] <- filtered_data[, metrics_modified] / (filtered_data$Min / 90)

```

We have "Tackles" and "Tackles Won" metrics, I want to create a "Tackle Success %" column

``` {r}

filtered_data["Tackles Won %"] <- filtered_data["Tackles Won"] / filtered_data["Tackles"]
```
